Follow-up Samples

### 2017-01-05 submission/manuscript_outline_20161024.tex
update bib, rename supplementa tables, and add Supplemental Figs component

Colorectal cancer (CRC) continues to be a worldwide health problem with
early detection being used as a key component in mitigating deaths due
to the disease. Previous research suggests a link between stool
bacterial microbiome and CRC. In this study, we used a model based on
the microbiome, demographics, and prior medical history to classify
individuals as having a lesion (i.e.~adenoma or carcinoma). We then used
this model to characterize the change in the gut microbiota before and
after surgery. The overall objective was to investigate the changes in
the microbiome after surgery in patients with adenomas or carcinomas.
This model was tested on a 66 person group that included samples before
and after treatment to allow for the assessment of how the model adjusts
risk after treatment. The model used for prediction had an AUC of 0.763.
For the follow up samples our Random Forest model significantly
decreased the positive probability of lesion compared to the initial
samples for both adenoma (P-value = 3.64e-11) and carcinoma (P-value =
7.95e-08). Our model predicted that 36.4\% of the 67-person cohort had
normal colons and 63.6\% had a lesion. Some OTUs that changed the most
before and after treatment included OTUs that were affiliated with
members of Blautia, Streptococcus and Escherichia/Shigella. Our model
suggests that treatment does significantly reduce the probability of
having a colonic lesion. Further surveillance of these individuals will
enable us to determine whether models such as the one we present here
can also be used to predict recurrence of colorectal cancer.

@@

Colorectal cancer (CRC) continues to be a leading cause of cancer
related deaths and is the second most common cancer death among men aged
40-79 years of age (1, 2). Over the last few years death due to the
disease has seen a significant decrease thanks mainly to improvements in
screening (1). However, despite this giant improvement there are still
approximately 50,000 deaths from the disease a year (2). It is estimated
that around 5-10\% of all CRCs can be explained by autosomal dominant
inheritance (3). The vast majority of CRCs are not inherited and the
exact etiology to disease has not been well worked out (2). Although
many risk factors have been identified ({\textbf{???}}) and non-invasive
screening techniques have started to be put into consistent use (4, 5)
there has been a consistent increase in the incidence of CRC in the
younger population.

@@

\textbf{\emph{Outcome of Model Training}} The range of the AUC for model
training ranged from a minimum of 0.723 to a maximum of 0.795 with the
middle of all 100 runs having an AUC of 0.761. Interestingly, the worst
AUC model from training performed the best on it's respective 80/20
split test data {[}Figure 3{]}. In fact the 80/20 test performance
showed that the AUC for the middle model chosen was the most stable
(best training model test set AUC = 0.646, middle training model test
set AUC = 0.744, worse training model test set AUC = 0.904). That is to
say it had the smallest change in AUC in comparison to the minimum and
maximum AUC trained models. The middle model was close to the full
training data AUC 0.763. There was no significant difference between the
AUC of the best and middle training models (P-value = 1). There was also
no difference in the middle model versus worse (P-value = 0.0431) or
full data model (P-value = 1). The two comparisons with a significant
difference were between the worse training model and the best training
model (P-value = 6.83e-04) and the worse training model and full
training model (P-value = 1.2e-03).

\textbf{\emph{Most Important Variables to the Model}} Overall, there
were a total of 37 variables identified as being present in more than
50\% of the training models {[}Table S2{]}. The top 5 most important
bacterial OTUs were Lachnospiraceae (Otu000013), Escherichia/Shigella
(Otu000018), Ruminococcaceae (Otu000020), Ruminococcus (Otu000017), and
Porphyromonas (Otu000153). These 5 OTUs were present in at least 90 out
of the total 100 different 80/20 runs.

@@

There was 1 individual who still clearly had CRC on follow up as well as
5 individuals whose status on follow up was unknown. Although the 1
individual had a decrease in positive probability their follow up sample
was still higher than the cutoff threshold of 0.5 (positive probability
= 0.903). Interestingly, 1 individuals who were unknown on follow up
still were over the threshold cutoff of 0.5 even though, like the 1
individual with clear CRC on follow up, the probability of an adenoma or
carcinoma decreased {[}Table S3{]}.

@@

\textbf{\emph{Specific OTUs in the Lesion Model are not Detected in
Follow Up Versus Initial Samples}} Overall, there were a total of 11
OTUs that were common between the main lesion model and the model for
classifying initial and follow up samples specifically {[}Table S4{]}. A
total of 1 OTU was still significant after multiple comparison
correction and it's lowest taxonomic identification was to Blautia. In
general, Otu000012 (Blautia) was decreased from initial to follow up
{[}Figure 5{]}. The relative abundance was not drastically different
then the mean of the values observed in the control training set
{[}Figure 5{]}.

\textbf{\emph{Differences in Adenoma and Carcinoma in Previously
Associated Cancer Bacteria}} First, there was a clear magnitude
difference in these specific OTUs based on whether they were from
adenoma or carcinoma individuals {[}Figure 6{]}. The carcinoma samples
showed a significant difference between initial and follow up samples
for Peptostreptococcus stomatis (P-value = 0.0183) and Porphyromonas
asaccharolytica (P-value = 0.0154) whereas there were no signficant
differences in any of these OTUs in the adenoma samples {[}Table S5{]}.

@@

We then investigated which OTUs could potentially be more important in
our model {[}Figure 5 \& Table S4{]}. Many of the OTUs identified
classified to normal flora bacterium {[}Table S4{]}. Only a single OTU
though was significant after multiple comparison correction and the
lowest taxonomic identification of Otu000012 was to Blautia. Although
there was a difference in the relative abundance at initial and follow
up these values were not drastically different from the relative
abundance values observed in the control individuals of the training set
{[}Figure 5{]}. Although we were interested in what we could use to
classify those with either adenoma or carcinoma versus normal. We found
that the traditional bacteria associated with CRC were higher in
magnitude in the carcinoma group and there were significant differences
in some of these OTUs between the initial and follow up samples
{[}Figure 5 and Tabl S5{]}. This research provides evidence that it is
possible to use bacterial microbiome data to create a highily sensitive
model, that is reactive to therapy, for detection of adenoma or
carcinoma. It accomplishes this by using a unique sample set in which
before and after surgery stool samples are available for assessment. By
using these types of samples we are not only able to show sensitivity of
lesion prediction but also able to show that this model is reactive.
That is to say that after surgery for removal of the adenoma or
carcinoma it decreases the positive probability to reflect a lower
likelihood of the individual having an adenoma or carcinoma.

@@

\textbf{\emph{Study Design and Patient Sampling}} The sampling and
design of the study was similar to that reported in Baxter, et al (13).
In brief, study exclusion involved those who had already undergone
surgery, radiation, or chemotherapy, had colorectal cancer before a
baseline stool sample could be obtained, had IBD, a known hereditary
non-polyposis colorectal cancer, or Familial adenomatous polyposis.
Samples used to build the model used for prediction were collected
either prior to a colonoscopy or between 1 - 2 weeks after. The
bacterial microbiome has been shown to nomralize within this time period
(\textbf{insert citation}). Kept apart from this training set were a
total of 67 individuals that not only had a sample as described
previoulsy but also a follow up sample between 188 - 546 days after
surgery and treatment had been completed. This study was approved by the
University of Michigan Institutional Review Board. All study
participants provided informed consent and the study itself conformed to
the guidelines set out by the Helsinki Declaration.

\textbf{\emph{Fecal Immunochemical Test and 16S rRNA Gene Sequencing}}
FIT was analyzed as previously published using both OC FIT-CHEK and
OC-Auto Micro 80 automated system (Polymedco Inc.) (\textbf{insert
citation}). 16S rRNA gene sequencing was completed as previously
described by Kozich, et al. (18). In brief, DNA extraction used the 96
well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075
automated pipetting system (Eppendorf). The V4 variable region was
amplified and the resulting product was split between three sequencing
runs with control, adenoma, and carcinoma evenly represented on each
run. Each group was randomly assigned to avoid biases based on sample
collection location.

\textbf{\emph{Sequence Processing}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences. This process
has been previously described (\textbf{insert citations}). The general
processing workflow using mothur is as follows: Paired-end reads were
first merged into contigs, quality filtered, aligned to the SILVA
database, screening for chimeras, classified with a naive Bayesian
classifier using the Ribosomal Database Project (RDP), and clustered
into Operational Taxonomic Units (OTUs) using a 97\% similarity cutoff
with an average neighbor clustering algorithm. The numer of sequences
for each sample was rarified to 10521 in an attempt to minimize uneven
sampling.

\textbf{\emph{Lesion Model Creation}} The Random Forest (19) algorithm
was used to create the model used for prediction of lesion (adenoma or
carcinoma) for the 67 individuals with follow up samples. The model
included data on FIT and the bacterial microbiome. Non-binary data was
checked for near zero variance and auto correlation. Data columns that
had near zero variance were removed. Columns that were correlated with
each other over a Spearman correlation coefficient of 0.75 had one of
the two columns removed. This pre-processing was performed with the R
package caret (v6.0.73). Optimization of the mtry hyperparameter
involved taking the samples and making 100 80/20 (train/test) splits in
the data where control and lesion were equally represented in the 80 and
20 split, respectively. This 80\% portion was then split again into an
80/20 split, and run through 20 repeated 10-fold cross validations to
optimzie the model's AUC (Area Under the Curve of the Receiver Operator
Characteristic). This resulting model was then tested on the 20\% of the
data that was originally held out from this overall process. Once the
ideal mtry was found the entire 490 sample set was used to create the
final Random Forest model on which testing on the 67-person cohort was
completed. The default cutoff of 0.5 was used as the threshold to
classify individuals as positive or negative for lesion. The
hyperparameter, mtry, defines the number of variables to investigate at
each split before a new division of the data is created.

@@

\textbf{\emph{Statistical Analysis}} The R software package (v3.3.0) was
used for all statisitical analysis. Comparisons between bacterial
community structure utilized PERMANOVA (\textbf{insert citation}) in the
vegan package (v2.4.1) while comparisons between ROC curves utilized the
method by DeLong et al. (\textbf{insert citation}) executed by the pROC
(v1.8) package. Comparisons between probabilities as well as overall
amount of OTU between initial and follow up samples utilized a paired
wilcoxson ranked sum test. Where multiple comparison testing was needed
a Benjamini-Hochberg (BH) correction was applied (\textbf{insert
citation}) and a corrected P-value of less than 0.05 was considered
significant. Unless otherwise stated the P-values reported are those of
the BH corrected ones.

@@

\textbf{Figure S1: Thetayc Graphed Against Time of Follow up Sample from
Initial}

@@

\hypertarget{ref-kozich_development_2013}{}
18. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq Illumina sequencing platform.
Applied and Environmental Microbiology \textbf{79}:5112--5120.
doi:\href{https://doi.org/10.1128/AEM.01043-13}{10.1128/AEM.01043-13}.

\hypertarget{ref-breiman_random_2001}{}
19. \textbf{Breiman L}. 2001. Random Forests. Machine Learning
\textbf{45}:5--32.
doi:\href{https://doi.org/10.1023/a:1010933404324}{10.1023/a:1010933404324}.

>>>>

Colorectal cancer (CRC) continues to be a worldwide health problem with
early detection being used as a key component in mitigating deaths due
to the disease. Previous research suggests a link between stool
bacterial microbiome and CRC. In this study, we used a model based on
the microbiome, demographics, and prior medical history to classify
individuals as having a lesion (i.e.~adenoma or carcinoma). We then used
this model to characterize the change in the gut microbiota before and
after surgery. The overall objective was to investigate the changes in
the microbiome after surgery in patients with adenomas or carcinomas.
This model was tested on a 66 person group that included samples before
and after treatment to allow for the assessment of how the model adjusts
risk after treatment. The model used for prediction had an AUC of NA.
For the follow up samples our Random Forest model significantly
decreased the positive probability of lesion compared to the initial
samples for both adenoma (P-value = 3.64e-11) and carcinoma (P-value =
7.95e-08). Our model predicted that 36.4\% of the 67-person cohort had
normal colons and 63.6\% had a lesion. Some OTUs that changed the most
before and after treatment included OTUs that were affiliated with
members of Blautia, Streptococcus and Escherichia/Shigella. Our model
suggests that treatment does significantly reduce the probability of
having a colonic lesion. Further surveillance of these individuals will
enable us to determine whether models such as the one we present here
can also be used to predict recurrence of colorectal cancer.

@@

Colorectal cancer (CRC) continues to be a leading cause of cancer
related deaths and is the second most common cancer death among men aged
40-79 years of age (1, 2). Over the last few years death due to the
disease has seen a significant decrease thanks mainly to improvements in
screening (1). However, despite this giant improvement there are still
approximately 50,000 deaths from the disease a year (2). It is estimated
that around 5-10\% of all CRCs can be explained by autosomal dominant
inheritance (3). The vast majority of CRCs are not inherited and the
exact etiology to disease has not been well worked out (2). Although
many risk factors have been identified (2) and non-invasive screening
techniques have started to be put into consistent use (4, 5) there has
been a consistent increase in the incidence of CRC in the younger
population.

@@

\textbf{\emph{Outcome of Model Training}} The range of the AUC for model
training ranged from a minimum of 0.723 to a maximum of 0.795 with the
middle of all 100 runs having an AUC of 0.761. Interestingly, the worst
AUC model from training performed the best on it's respective 80/20
split test data {[}Figure 3{]}. In fact the 80/20 test performance
showed that the AUC for the middle model chosen was the most stable
(best training model test set AUC = 0.646, middle training model test
set AUC = 0.744, worse training model test set AUC = 0.904). That is to
say it had the smallest change in AUC in comparison to the minimum and
maximum AUC trained models. The middle model was close to the full
training data AUC NA. There was no significant difference between the
AUC of the best and middle training models (P-value = 1). There was also
no difference in the middle model versus worse (P-value = 0.0431) or
full data model (P-value = 1). The two comparisons with a significant
difference were between the worse training model and the best training
model (P-value = 6.83e-04) and the worse training model and full
training model (P-value = 1.2e-03).

\textbf{\emph{Most Important Variables to the Model}} Overall, there
were a total of 37 variables identified as being present in more than
50\% of the training models {[}Figure S2{]}. The top 5 most important
bacterial OTUs were Lachnospiraceae (Otu000013), Escherichia/Shigella
(Otu000018), Ruminococcaceae (Otu000020), Ruminococcus (Otu000017), and
Porphyromonas (Otu000153). These 5 OTUs were present in at least 90 out
of the total 100 different 80/20 runs.

@@

There was 1 individual who still clearly had CRC on follow up as well as
5 individuals whose status on follow up was unknown. Although the 1
individual had a decrease in positive probability their follow up sample
was still higher than the cutoff threshold of 0.5 (positive probability
= 0.903). Interestingly, 1 individuals who were unknown on follow up
still were over the threshold cutoff of 0.5 even though, like the 1
individual with clear CRC on follow up, the probability of an adenoma or
carcinoma decreased {[}Table S2{]}.

@@

\textbf{\emph{Specific OTUs in the Lesion Model are not Detected in
Follow Up Versus Initial Samples}} Overall, there were a total of 11
OTUs that were common between the main lesion model and the model for
classifying initial and follow up samples specifically {[}Table S3{]}. A
total of 1 OTU was still significant after multiple comparison
correction and it's lowest taxonomic identification was to Blautia. In
general, Otu000012 (Blautia) was decreased from initial to follow up
{[}Figure 5{]}. The relative abundance was not drastically different
then the mean of the values observed in the control training set
{[}Figure 5{]}.

\textbf{\emph{Differences in Adenoma and Carcinoma in Previously
Associated Cancer Bacteria}} First, there was a clear magnitude
difference in these specific OTUs based on whether they were from
adenoma or carcinoma individuals {[}Figure 6{]}. The carcinoma samples
showed a significant difference between initial and follow up samples
for Peptostreptococcus stomatis (P-value = 0.0183) and Porphyromonas
asaccharolytica (P-value = 0.0154) whereas there were no signficant
differences in any of these OTUs in the adenoma samples {[}Table S4{]}.

@@

We then investigated which OTUs could potentially be more important in
our model {[}Figure 5 \& Table S3{]}. Many of the OTUs identified
classified to normal flora bacterium {[}Table S3{]}. Only a single OTU
though was significant after multiple comparison correction and the
lowest taxonomic identification of Otu000012 was to Blautia. Although
there was a difference in the relative abundance at initial and follow
up these values were not drastically different from the relative
abundance values observed in the control individuals of the training set
{[}Figure 5{]}. Although we were interested in what we could use to
classify those with either adenoma or carcinoma versus normal. We found
that the traditional bacteria associated with CRC were higher in
magnitude in the carcinoma group and there were significant differences
in some of these OTUs between the initial and follow up samples
{[}Figure 5 and Tabl S5{]}. This research provides evidence that it is
possible to use bacterial microbiome data to create a highily sensitive
model, that is reactive to therapy, for detection of adenoma or
carcinoma. It accomplishes this by using a unique sample set in which
before and after surgery stool samples are available for assessment. By
using these types of samples we are not only able to show sensitivity of
lesion prediction but also able to show that this model is reactive.
That is to say that after surgery for removal of the adenoma or
carcinoma it decreases the positive probability to reflect a lower
likelihood of the individual having an adenoma or carcinoma.

@@

\textbf{\emph{Study Design and Patient Sampling}} The sampling and
design of the study was similar to that reported in Baxter, et al (13).
In brief, study exclusion involved those who had already undergone
surgery, radiation, or chemotherapy, had colorectal cancer before a
baseline stool sample could be obtained, had IBD, a known hereditary
non-polyposis colorectal cancer, or Familial adenomatous polyposis.
Samples used to build the model used for prediction were collected
either prior to a colonoscopy or between 1 - 2 weeks after. The
bacterial microbiome has been shown to nomralize within this time period
(18). Kept apart from this training set were a total of 67 individuals
that not only had a sample as described previoulsy but also a follow up
sample between 188 - 546 days after surgery and treatment had been
completed. This study was approved by the University of Michigan
Institutional Review Board. All study participants provided informed
consent and the study itself conformed to the guidelines set out by the
Helsinki Declaration.

\textbf{\emph{Fecal Immunochemical Test and 16S rRNA Gene Sequencing}}
FIT was analyzed as previously published using both OC FIT-CHEK and
OC-Auto Micro 80 automated system (Polymedco Inc.) (19). 16S rRNA gene
sequencing was completed as previously described by Kozich, et al. (20).
In brief, DNA extraction used the 96 well Soil DNA isolation kit (MO BIO
Laboratories) and an epMotion 5075 automated pipetting system
(Eppendorf). The V4 variable region was amplified and the resulting
product was split between three sequencing runs with control, adenoma,
and carcinoma evenly represented on each run. Each group was randomly
assigned to avoid biases based on sample collection location.

\textbf{\emph{Sequence Processing}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences. This process
has been previously described (20). The general processing workflow
using mothur is as follows: Paired-end reads were first merged into
contigs, quality filtered, aligned to the SILVA database, screening for
chimeras, classified with a naive Bayesian classifier using the
Ribosomal Database Project (RDP), and clustered into Operational
Taxonomic Units (OTUs) using a 97\% similarity cutoff with an average
neighbor clustering algorithm. The numer of sequences for each sample
was rarified to 10521 in an attempt to minimize uneven sampling.

\textbf{\emph{Lesion Model Creation}} The Random Forest (21) algorithm
was used to create the model used for prediction of lesion (adenoma or
carcinoma) for the 67 individuals with follow up samples. The model
included data on FIT and the bacterial microbiome. Non-binary data was
checked for near zero variance and auto correlation. Data columns that
had near zero variance were removed. Columns that were correlated with
each other over a Spearman correlation coefficient of 0.75 had one of
the two columns removed. This pre-processing was performed with the R
package caret (v6.0.73). Optimization of the mtry hyperparameter
involved taking the samples and making 100 80/20 (train/test) splits in
the data where control and lesion were equally represented in the 80 and
20 split, respectively. This 80\% portion was then split again into an
80/20 split, and run through 20 repeated 10-fold cross validations to
optimzie the model's AUC (Area Under the Curve of the Receiver Operator
Characteristic). This resulting model was then tested on the 20\% of the
data that was originally held out from this overall process. Once the
ideal mtry was found the entire 490 sample set was used to create the
final Random Forest model on which testing on the 67-person cohort was
completed. The default cutoff of 0.5 was used as the threshold to
classify individuals as positive or negative for lesion. The
hyperparameter, mtry, defines the number of variables to investigate at
each split before a new division of the data is created.

@@

\textbf{\emph{Statistical Analysis}} The R software package (v3.3.0) was
used for all statisitical analysis. Comparisons between bacterial
community structure utilized PERMANOVA (22) in the vegan package
(v2.4.1) while comparisons between ROC curves utilized the method by
DeLong et al. (23) executed by the pROC (v1.8) package. Comparisons
between probabilities as well as overall amount of OTU between initial
and follow up samples utilized a paired wilcoxson ranked sum test. Where
multiple comparison testing was needed a Benjamini-Hochberg (BH)
correction was applied (24) and a corrected P-value of less than 0.05
was considered significant. Unless otherwise stated the P-values
reported are those of the BH corrected ones.

@@

\textbf{Figure S1: Distribution of P-values from Paired Wilcoxson
Analysis of OTUs in Initial versus Follow Up}

\textbf{Figure S2: Summary of Important Variables in both Lesion and
Initial/Follow Up models} A) MDA of the most important variables in the
lesion model. The black point represents the median and the different
colors are the different runs up to 100. B) The total number of
appearances of each variable in the 100 different lesion models. The
cutoff of 50\% was used to assess importance.

\textbf{Figure S3: Thetayc Graphed Against Time of Follow up Sample from
Initial}

@@

\hypertarget{ref-obrien_impact_2013}{}
18. \textbf{O'Brien CL}, \textbf{Allison GE}, \textbf{Grimpen F},
\textbf{Pavli P}. 2013. Impact of colonoscopy bowel preparation on
intestinal microbiota. PloS One \textbf{8}:e62815.
doi:\href{https://doi.org/10.1371/journal.pone.0062815}{10.1371/journal.pone.0062815}.

\hypertarget{ref-daly_evaluation_2013}{}
19. \textbf{Daly JM}, \textbf{Bay CP}, \textbf{Levy BT}. 2013.
Evaluation of fecal immunochemical tests for colorectal cancer
screening. Journal of Primary Care \& Community Health
\textbf{4}:245--250.
doi:\href{https://doi.org/10.1177/2150131913487561}{10.1177/2150131913487561}.

\hypertarget{ref-kozich_development_2013}{}
20. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq Illumina sequencing platform.
Applied and Environmental Microbiology \textbf{79}:5112--5120.
doi:\href{https://doi.org/10.1128/AEM.01043-13}{10.1128/AEM.01043-13}.

\hypertarget{ref-breiman_random_2001}{}
21. \textbf{Breiman L}. 2001. Random Forests. Machine Learning
\textbf{45}:5--32.
doi:\href{https://doi.org/10.1023/a:1010933404324}{10.1023/a:1010933404324}.

\hypertarget{ref-anderson_permanova_2013}{}
22. \textbf{Anderson MJ}, \textbf{Walsh DCI}. 2013. PERMANOVA, ANOSIM,
and the Mantel test in the face of heterogeneous dispersions: What null
hypothesis are you testing? Ecological Monographs \textbf{83}:557--574.
doi:\href{https://doi.org/10.1890/12-2010.1}{10.1890/12-2010.1}.

\hypertarget{ref-delong_comparing_1988}{}
23. \textbf{DeLong ER}, \textbf{DeLong DM}, \textbf{Clarke-Pearson DL}.
1988. Comparing the areas under two or more correlated receiver
operating characteristic curves: A nonparametric approach. Biometrics
\textbf{44}:837--845.

\hypertarget{ref-benjamini_controlling_1995}{}
24. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
false discovery rate: A practical and powerful approach to multiple
testing. Journal of the Royal Statistical Society Series B
(Methodological) \textbf{57}:289--300.
